Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-07-19 submission/Sze_Obesity_mBio_2016.Rmd
Add expanded Random Forest analysis; Closes Issue #18

***Development of a microbiome-based classifier of obesity.***
The Walters re-analysis study suggested that it was possible to classify individuals as being non-obese or obese  based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The cross-validated AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this probability to define a threshold for calculating the accuracy of the models when applied to the other `r number_string[n-1]` datasets (Figure 5). Although there was considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

@@

***Power and Sample Size Estimate Simulations.***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S4-S9). Although there is no biological rationale for these effect sizes, they represent a range that includes effect sizes that would be generally considered to be biologically significant. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S4). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median required sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `r signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

@@

Similar to our study, the Walters study generated Random Forest machine learning models to differentiate between non-obese and obese individuals [@walters_meta-analyses_2014]. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.

@@

**Figure S4: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S5: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S6: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S8: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S9: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

>>>>

***Development of a microbiome-based classifier of obesity.***
The Walters re-analysis study suggested that it was possible to classify individuals as being non-obese or obese  based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values for the OTU-based models varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). To test models on other datasets, we trained models using genus-level phylotype data for each dataset. The cross-validated AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this probability to define a threshold for calculating the accuracy of the models when applied to the other `r number_string[n-1]` datasets (Figure 5). Although there was considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). We built similar models using taxonomic representation based on phylum, class, order, and family assignments and saw no variation in the results (Figure S4). We also attempted to predict individual BMI values as continuous variables based on the relative abundance of OTUs and genera. The median percent of the variance explained with the resulting models was `r median(otu$regression_auc, na.rm=T)` for the OTU-based models and `r median(genus$regression_auc, na.rm=T)` for the genus-based models. When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using relative abundance data from the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

@@

***Power and Sample Size Estimate Simulations.***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S5-S10). Although there is no biological rationale for these effect sizes, they represent a range that includes effect sizes that would be generally considered to be biologically significant. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median required sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `r signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

@@

Similar to our study, the Walters study generated Random Forest machine learning models to differentiate between non-obese and obese individuals [@walters_meta-analyses_2014]. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Furthermore, others have reported improved classification at broader taxonomic levels [@Sun2010]; we did not find this to be the case across the studies in our analysis (Figure S4). Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and that there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals to a BMI category based on their microbiota.

@@

**Figure S4: Overall accuracy of each study to predict non-obese and obese individuals based on that study's Random Forest machine learning model applied to each of the other studies when trained using relative abundance of each phylum, class, order, family, or genus.** The cross-validated AUC values for the training model are provided for each study and taxonomic level.

**Figure S5: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S6: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S7: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S8: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S9: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S10: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
